---
title: Using Large Language Models for Data Collection/Annotation in Social Sciences and Humanities
jupyter: ir
---



```{r}
#| vscode: {languageId: r}
library(ellmer)    # LLM API interfacing in R
library(irr)       # inter-rater reliability (Krippendorff's Alpha)
library(cli)       # progress bars
library(jsonlite)  # JSONL logging
library(tidyverse) # data manipulation and visualization
```

## Data loading

We encourage you to use the datasets available in SANE. The default code reuses the toy dataset from the previous notebook. 

```{r}
#| vscode: {languageId: r}
# Load CSV into a dataframe
data_url <- "../data/srl_data_example.csv"
df <- read_csv(data_url)
```

Note that only the first 10 rows contain the anonymized text of the conversations. We will use these texts for the prompting experiments in this notebook.

```{r}
#| vscode: {languageId: r}
# Use these 10 rows to define test_ids and test_conversations
test_ids <- df |> slice(1:10) |> pull(id) |> as.character()
test_conversations <- df |> slice(1:10) |> pull(conversation)
```

## Local open-weights deployment in SANE with Ollama

We will use Ollama to run open-weights models locally. 
Ollama is installed by default in the SANE environment, but on your own machine, you will need to pull (i.e., download) a model and start the local server to use it.

Example commands (run in a command line terminal):

- `ollama list` (to see available models)
- `ollama serve`  (starts the local server at http://localhost:11434; skip if already running)
- `ollama run <model_name>` (e.g., `ollama run qwen2.5:7b` to run the 7B version of the Qwen2.5 model)

Currently, we have the following models available locally via Ollama:
- qwen2.5:7b
- qwen2.5:14b
- qwen2.5-coder:7b
- qwen2.5-coder:14b
- gpt-oss:20b

The number after the colon indicates the number of parameters in the model (e.g., `qwen2.5:7b` has 7 billion parameters). Larger numbers indicate more parameters, hence bigger and more powerful (but often slower) models. The "coder" models are optimized for code generation, which may be useful for coding support and debugging.

```{r}
#| vscode: {languageId: r}
# Set up your model
model_name <- "qwen2.5:7b"
temperature <- 0  
max_tokens <- 1000
seed <- 123

# Helper: create a fresh chat object for each request
# To prompt a self-hosted Ollama model, we simply point ellmer to the local server.

make_chat <- function(system_prompt = NULL) {
  api_args <- list(temperature = temperature, max_tokens = max_tokens)
  if (!is.null(seed)) api_args$seed <- seed
  chat_ollama(
    model         = model_name,
    system_prompt = system_prompt,
    api_args      = api_args,
    echo          = "none",
    base_url      = "http://localhost:11434"
  )
}
```

## Working with a single prompt

Let's start with the system prompt (i.e., high-level instruction to the model).

```{r}
#| vscode: {languageId: r}
# Define a system prompt that explains the task and scoring rubric
system_prompt <- "
You are an expert in educational assessment and goal evaluation, with
specialized expertise in applying deductive coding schemes to score the quality
and content of student goals.

##TASK##
A university student was given a series of prompts, guiding them through the
process of setting and elaborating on an academic goal for the coming week. You
will be provided with the entire conversation including the prompts, and the
student answers. Your objective is to assess the specificity of of the student's
goal on a scale of 0 to 2 based on the entire conversation.
"
```

In `ellmer`, a fresh chat object carries the system prompt and the user's conversation text is passed directly to `chat$chat()`. Prompt the model and inspect the response!

```{r}
#| vscode: {languageId: r}
# Create a chat and make the API call to get a single response
chat <- make_chat(system_prompt = system_prompt)
single_response <- chat$chat(test_conversations[[1]])
cat(single_response)
```

Voila! You have your first successful prompting interaction with a local LLM API!

## Working with multiple prompts
Next, we go beyond a single prompt. Instead, we will work with **multiple prompts** at the same time.

**Tip**: Start with a small number of rows first to estimate time and (if using a paid API) cost to avoid surprises.

```{r}
#| vscode: {languageId: r}
# Iterate over the conversations and score them
multiple_responses <- list()
cli_progress_bar("Processing Requests", total = length(test_ids))
for (i in seq_along(test_ids)) {
  chat <- make_chat(system_prompt = system_prompt)
  multiple_responses[[test_ids[[i]]]] <- chat$chat(test_conversations[[i]])
  cli_progress_update()
}
cli_progress_done()
```

Inspect the responses!

```{r}
#| vscode: {languageId: r}
# Show an example response
cat(multiple_responses[["chat_2"]])
```

## Using structured output with a single prompt

Use the `type_object()` and related `type_*()` functions from `ellmer` to specify the desired output format. Unlike the Python/LangChain approach — where providers like Hugging Face require explicit JSON instructions and manual parsing — `ellmer`'s `$chat_structured()` method works uniformly across providers by leveraging each provider's tool-calling or structured output API under the hood.

For example:

```{r}
#| vscode: {languageId: r}
# Define the expected structured output schema
schema <- list(
  type = "object",
  additionalProperties = FALSE,
  required = c("goal_specificity", "reasoning"),
  properties = list(
    goal_specificity = list(
      type = "integer",
      description = "Score for goal specificity. Only return an integer from 0 to 2",
      minimum = 0,
      maximum = 2
    ),
    reasoning = list(
      type = "string",
      description = "The reasoning to justify the score"
    )
  )
)

output_structure <- type_from_schema(
  jsonlite::toJSON(schema, auto_unbox = TRUE)
)
```

Try with a single prompt request.

```{r}
#| vscode: {languageId: r}
# Create a chat and get a structured response for the first conversation
chat <- make_chat(system_prompt = system_prompt)
single_structured_response <- chat$chat_structured(test_conversations[[1]], type = output_structure)

# Convert to a plain list for inspection
as.list(single_structured_response)
```

## Using structured output with multiple prompts

Being able to work with multiple prompts at the same time and obtain structured output will save you a substantial amount of time in research projects!

```{r}
#| vscode: {languageId: r}
# Score multiple conversations with structured output
multiple_structured_responses <- list()
cli_progress_bar("Processing Messages", total = length(test_ids))
for (i in seq_along(test_ids)) {
  chat <- make_chat(system_prompt = system_prompt)
  multiple_structured_responses[[test_ids[[i]]]] <- chat$chat_structured(
    test_conversations[[i]], type = output_structure
  )
  cli_progress_update()
}
cli_progress_done()
```

Display all the structured responses:

```{r}
#| vscode: {languageId: r}
# Extract the scores from the structured responses
structured_scores <- map_int(multiple_structured_responses, "goal_specificity")
structured_scores
```

```{r}
#| vscode: {languageId: r}
# Extract the reasoning text from the structured responses
structured_reasonings <- map_chr(multiple_structured_responses, "reasoning")
structured_reasonings
```

## Enhancing reproducibility: log prompts and decisions

To make your data collection/annotation reproducible, log each prompt, model configuration, and model output. The block below writes CSV and JSONL logs to the `logs/` folder with a timestamped filename.

As these logs include prompts and model outputs, make sure to remove or anonymize sensitive information before sharing logs.

We show below how to log the prompts and responses from the structured output experiment, but you can apply the same logic to log any other prompting experiment you do in this notebook or in your own projects.

```{r}
#| vscode: {languageId: r}
# Build a prompt + decision log
dir.create("logs", showWarnings = FALSE)
timestamp     <- format(Sys.time(), "%Y%m%d_%H%M%S")
log_csv_path  <- paste0("logs/prompt_log_", timestamp, ".csv")
log_jsonl_path <- paste0("logs/prompt_log_", timestamp, ".jsonl")

# Score multiple conversations with structured output
multiple_structured_responses <- list()  # Store the structured responses
prompt_logs                   <- list()  # Store the prompt logs

cli_progress_bar("Processing Messages", total = length(test_ids))
for (i in seq_along(test_ids)) {
  chat <- make_chat(system_prompt = system_prompt)
  structured_response <- chat$chat_structured(test_conversations[[i]], type = output_structure)
  multiple_structured_responses[[test_ids[[i]]]] <- structured_response

  # Log the prompt, model config, and output
  prompt_logs[[i]] <- list(
    request_id    = test_ids[[i]],
    provider      = default_provider,
    model         = model_name,
    temperature   = temperature,
    max_tokens    = max_tokens,
    seed          = if (is.null(seed)) NA else seed,
    system_prompt = trimws(system_prompt),
    user_prompt   = test_conversations[[i]],
    score         = structured_response$goal_specificity,
    reasoning     = structured_response$reasoning
  )
  cli_progress_update()
}
cli_progress_done()

# Save logs to CSV and JSONL
prompt_log_df <- bind_rows(prompt_logs)
write_csv(prompt_log_df, log_csv_path)
# Write each record as a separate JSON line (JSONL format)
write_lines(
  sapply(prompt_logs, \(x) toJSON(x, auto_unbox = TRUE)),
  log_jsonl_path
)

# View the prompt log dataframe
head(prompt_log_df, 3)
```

## Check annotation quality

Implement a handy function to calculate Krippendorff's Alpha (i.e., agreement) between two vectors of specificity scores.

```{r}
#| vscode: {languageId: r}
compute_krippendorff_alpha <- function(x, y) {
  # Format data into a reliability matrix (rows = raters, cols = items)
  rating_matrix <- rbind(x, y)
  # Compute Krippendorff's Alpha (ordinal metric)
  kripp.alpha(rating_matrix, method = "ordinal")
}
```

Let's check the agreement between the specificity scores we got from the LLM above and the human expert-coded specificity scores!

```{r}
#| vscode: {languageId: r}
# Compare agreement between expert and LLM ratings
expert_specificity_scores         <- df |> slice(1:10) |> pull(expert_specificity_score)
structured_llm_specificity_scores <- map_int(multiple_structured_responses, "goal_specificity")
cat("Krippendorff's Alpha:")
compute_krippendorff_alpha(structured_llm_specificity_scores, expert_specificity_scores)
```

Not a great agreement score!

How about the agreement between the LLM specificity scores that already came with the dataset (i.e., column `best_llm_specificity_score`) and the human expert-coded scores?

Note that `best_llm_specificity_score` is based on prompts that were carefully engineered by Gabrielle.

```{r}
#| vscode: {languageId: r}
best_llm_specificity_scores <- df |> slice(1:10) |> pull(best_llm_specificity_score)
cat("Krippendorff's Alpha:")
compute_krippendorff_alpha(best_llm_specificity_scores, expert_specificity_scores)
```

Wow! Much better!

## Exercise: Try a different dataset of choice!

